The Key to a Performant Data Parallel Compiler

### 2016-11-15 2017-spring-pldi/hsu-2017-key_data_parallel_compiler.tex
We have to use LaTeX I guess.

\documentclass[numbers,9pt]{sigplanconf}
% \usepackage[utf8]{inputenc}
\usepackage{array}
\usepackage{fontspec}
\setmainfont[Ligatures=TeX]{STIX}
\setmonofont{APL385 Unicode}
\usepackage[scaled]{helvet} % see www.ctan.org/get/macros/latex/required/psnfss/psnfss2e.pdf
\usepackage{url}                  % format URLs
\usepackage{listings}          % format code
\usepackage{enumitem}      % adjust spacing in enums
\usepackage[colorlinks=false,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}   % print a hyperlinked DOI
\usepackage{booktabs}

@@

\title{The Key to a Data Parallel Compiler}

\toappear{Permission to make digital or hard copies of all or part of this
work for personal or classroom use is granted without fee provided that copies
are not made or distributed for profit or commercial advantage and that copies
bear this notice and the full citation on the first page. Copyrights for
components of this work owned by others than the author(s) must be honored.
Abstracting with credit is permitted. To copy otherwise, or republish, to post
on servers or to redistribute to lists, requires prior specific permission
and/or a fee. Request permissions from Permissions@acm.org.

\emph{ARRAY'16}, June 13-17 2016, Santa Barbara, CA, USA

Copyright is held by the owner/author(s). Publication rights licensed to ACM.

ACM 978-1-4503-4384-8/16/06…\$15.00

DOI: http://dx.doi.org/10.1145/2935323.2935331}

\authorinfo{Aaron W. Hsu}{Indiana University, USA}{awhsu@indiana.edu}

\maketitle
\begin{abstract}
We present a language-driven strategy for the construction of 
compilers that are inherently data-parallel in their design and 
implementation. Using an encoding of the inter-node relationships 
between nodes in an AST called a Node Coordinate Matrix, we demonstrate how 
an operator called the Key operator, that applies a function over groupings 
of array cells grouped and ordered by their keys, when used in conjunction 
with a Node Coordinate Matrix, permits arbitrary computation over sub-trees 
of an AST using purely data-parallel array programming techniques. We discuss 
the use of these techniques when applied to an experimental commercial 
compiler called Co-dfns, which is a fully data-parallel compiler developed 
using the techniques discussed here.
\end{abstract}

\category{D.3.4}{Programming Languages}{Processors}[Compilers]

\keywords
APL, Parallel, Compilers, Key, Node Coordinates

% Proposed new outline:
% 
% I. 	Generalization on manipulations over Sub-tree computation
%   A.	Key + Encoding → Data parallel, Array-based computation
%   B.	Easy to write/read
%   C.	Uniform across passes
% II.	Case Studies
%   A.	Function Lifting
%   B.	Expression Flattening
%   C.	Lexical Scope Resolution
%   D.	Illustrates the problem space
% III.	Other passes/Issues
%   A.	Discussion of other passes
%   B.	Beyond the Key Operator
% IV.	In practice/Co-dfns Compiler
% V.	Performance/Results
% 
% Questions that must be addressed:
% 
% 1. Why the formalization in APL?
% 2. Why parallelize a compiler?
% 3. How well does it work/scale/generalize

@@

Compilers represent a peculiar class of tree/graph algorithms that greatly 
alter the underlying structure of the input graph into something very 
different in structure, but equivalent by some measure of interpretation. 
These algorithms are widely applicable, but have stubbornly resisted a 
general approach to parallelization. Most compilers are single-threaded 
or make very limited use of parallelism in an ad hoc way. The design and 
analysis of compilers almost universally deals with compilers as recursive 
traversals over ASTs. Such formulations usually include heavy reliance on 
branching, recursion, and sometimes very sophisticated control flow in 
practice. All of this contributes to make it difficult to effectively 
parallelize such algorithms onto architectures that are sensitive to 
branching and recursion, such as GPGPUs or highly vectorized CPUs.

@@

\begin{verbatim}
      1+⌈⌿Fd
5
      1+⌈⌿Ed
3
\end{verbatim}

@@

\begin{verbatim}
      ⍳1+⌈⌿Fd
0 1 2 3 4 5
      ⍳1+⌈⌿Ed
0 1 2
\end{verbatim}

@@

\begin{verbatim}
      (⍳3)∘.×⍳3
0 0 0
0 1 2
0 2 4
\end{verbatim}

@@

\begin{verbatim}
      Fd∘.=⍳1+⌈⌿Fd
1 0 0 0 0
0 1 0 0 0
0 0 1 0 0
0 0 0 1 0
0 0 0 0 1
0 0 0 0 1
0 0 0 0 1
0 0 1 0 0
0 0 0 1 0
      Ed∘.=⍳1+⌈⌿Ed
1 0 0
0 1 0
0 0 1
0 0 1
0 0 1
0 1 0
0 1 0
0 0 1
0 0 1
0 0 1
\end{verbatim}

>>>>

\documentclass[numbers,10pt,preprint]{sigplanconf}
\usepackage{array}
\usepackage{fontspec}
\setmainfont[Ligatures=TeX]{Times New Roman}
\setmonofont{APL385 Unicode}
\usepackage[scaled]{helvet} % see www.ctan.org/get/macros/latex/required/psnfss/psnfss2e.pdf
\usepackage{url}                  % format URLs
\usepackage{listings}          % format code
\usepackage{enumitem}      % adjust spacing in enums
\usepackage[colorlinks=false,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}   % print a hyperlinked DOI
\usepackage{booktabs}

@@

\title{The Key to a Performant Data Parallel Compiler}
\authorinfo{Aaron W. Hsu}{Indiana University, USA}{awhsu@indiana.edu}
\maketitle

\begin{abstract}
We present a language-driven strategy for the construction of compilers that are inherently data-parallel in their design and implementation. Using an encoding of the inter-node relationships between nodes in an AST called a Node Co-ordinate Matrix, we demonstrate how an operator called the Key operator, that applies a function over groupings of array cells grouped and ordered by their keys, when used in conjunction with a Node Coordinate Matrix, permits arbi-trary computation over sub-trees of an AST using purely data-parallel array programming techniques. We discuss the use of these techniques when applied to an experi-mental commercial compiler called Co-dfns, which is a fully data-parallel compiler developed using the techniques discussed here.
\end{abstract}

\category{D.3.4}{Programming Languages}{Processors}[Compilers]
\terms Algorithms, Performance, Design, Languages
\keywords APL, Parallel, Compilers, Key, Node Coordinates

@@

Compilers represent a peculiar class of tree/graph algo-rithms that greatly alter the underlying structure of the input graph into something very different in structure, but equivalent by some measure of interpretation. These algo-rithms are widely applicable, but have stubbornly resisted a general approach to parallelization. Most compilers are single-threaded or make very limited use of parallelism in an ad hoc way. The design and analysis of compilers al-most universally deals with compilers as recursive travers-als over ASTs. Such formulations usually include heavy reliance on branching, recursion, and sometimes very so-phisticated control flow in practice. All of this contributes to make it difficult to effectively parallelize such algo-rithms onto architectures that are sensitive to branching and recursion, such as GPGPUs or highly vectorised CPUs.

@@

\begin{verbatim}
      1+⌈⌿Fd  │        1+⌈⌿Ed
5             │  3
\end{verbatim}

@@

\begin{verbatim}
      ⍳1+⌈⌿Fd │        ⍳1+⌈⌿Ed
0 1 2 3 4 5   │  0 1 2
\end{verbatim}

@@

\begin{verbatim}
      (⍳3)∘.×⍳3  │        ⍳3
0 0 0            │  0 1 2
0 1 2            │        ⍴⍳3
0 2 4            │  3
\end{verbatim}

@@

\begin{verbatim}
      Fd∘.=⍳1+⌈⌿Fd │      Ed∘.=⍳1+⌈⌿Ed
1 0 0 0 0          │ 1 0 0
0 1 0 0 0          │ 0 1 0
0 0 1 0 0          │ 0 0 1
0 0 0 1 0          │ 0 0 1
0 0 0 0 1          │ 0 0 1
0 0 0 0 1          │ 0 1 0
0 0 0 0 1          │ 0 1 0
0 0 1 0 0          │ 0 0 1
0 0 0 1 0          │ 0 0 1
                   │ 0 0 1
\end{verbatim}
